---
title: 'Data607 Assignment 4: Tidying and Transforming Data'
author: "Anthony Josue Roman"
date: "2024-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(readr)
library(RCurl)
library(stringr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
```

This assignment will use the following packages:

```{r eval=FALSE}
library(readr)
library(RCurl)
library(stringr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
```

## General Overview

The goal of this assignment is to utilize untidy airport data that were on time or delayed from 5 different locations (cities) from two different airlines. The following data will also be uploaded onto my github [page](https://github.com/spacerome/Data607_Assignment_4). 

## Initilization

To get the data, we will extract the data from the github file from [here](https://raw.githubusercontent.com/spacerome/Data607_Assignment_4/refs/heads/main/airline_data.csv).

```{r get data}
raw_file <- getURL("https://raw.githubusercontent.com/spacerome/Data607_Assignment_4/refs/heads/main/airline_data.csv")

csvfile <- data.frame(read.csv(text=raw_file, sep= "\t", stringsAsFactors = FALSE, check.names = FALSE))
csvfile
```

Looking at the data the raw data gives `X` and `X.1` for the first columns which should be `airline` and `status` which should be fixed. The third and fifth rows have an empty value which will be fixed in the next section.

## Tidying the Data

```{r tidying data}
colnames(csvfile)[1] <- "airline"
colnames(csvfile)[2] <- "status"
csvfile
```

```{r include=FALSE}
# Alternate if the first one fails Uncomment if fails
# csvfile$airline[csvfile$airline == "    " | is.na(csvfile$airline)] <- c("ALASKA", "AM WEST")
# colnames(csvfile) <- gsub("\\.", " ", colnames(csvfile))
# csvfile
```

After fixing the columns, we will now fix the empty rows from rows 3 and 5, to have `ALASKA` and `AMWEST` to continue prepping the data for analysis on the next section. The cities that have a space (e.g. Los Angeles) were outputed as `Los.Angeles` which will be fixed by utilizing `gsub` to fix it.

```{r fixing empties}
csvfile[csvfile=="    "] <- NA
colnames(csvfile) <- gsub("\\.", " ", colnames(csvfile))
csvfile <- fill(csvfile, airline)
csvfile
```

After cleaning the data then we utilize `pivot_longer` to prep `csvfile` for further data analysis.

```{r make tables}
csvfile <- pivot_longer(csvfile, cols = !c(airline, status),
                        names_to = "city", values_to = "flights")
head(csvfile)
```

## Analyzing the Data

After cleaning the data we will have three variables utilized for analysis: `airlineData`, `cityData`, and `cityairlineData`.

```{r airline analysis}
airlineData <- csvfile %>%
  group_by(airline) %>% summarise(flightFrequency = sum(flights), flightDelay = sum(ifelse(status =='delayed',flights,0)))
airlineData <- airlineData %>%
  mutate(ontimeFlights = flightFrequency - flightDelay,
         delayRate = flightDelay/ontimeFlights,
         ontimeRate = ontimeFlights/flightFrequency) %>% arrange(delayRate)
airlineData
```

From analyzing the `airlineData`, `AM WEST` has the highest ontime rate of `89.1%`, and lowest delay rate of `12.2%`, whereas `ALASKA` has a higher delay rate of `15.3%`, but lower ontime rate of `86.7%`.

```{r city analysis}
cityData <- csvfile %>%
  group_by(city) %>% summarise(flightFrequency = sum(flights), flightDelay = sum(ifelse(status =='delayed',flights,0)))
cityData <- cityData %>%
  mutate(ontimeFlights = flightFrequency - flightDelay,
         delayRate = flightDelay/ontimeFlights,
         ontimeRate = ontimeFlights/flightFrequency) %>% arrange(delayRate)
cityData
```

From analyzing the `cityData`, Phoenix has the highest ontime Rate of `92.2%`, and lowest delay rate of `8.44%`. San Francisco has the highest delay rate of `28.1%` and its ontime rate is `78.1%`. 

```{r city airline analysis}
cityairlineData <- csvfile %>%
  group_by(city, airline) %>%
  summarise(flightFrequency = sum(flights, na.rm = TRUE), 
            flightDelay = sum(ifelse(status == 'delayed', flights, 0), na.rm = TRUE)) %>%
  mutate(ontimeFlights = flightFrequency - flightDelay,
         delayRate = flightDelay / flightFrequency) %>%
  ungroup()
cityairlineData
```

From this data, `AM WEST` had a higher number of flight delays than `ALASKA` within 4 of the 5 cities, but `ALASKA` had the highest delays at Seattle and had lesser delays on the other locations, but the delay rate for `AMWEST` is higher than `ALASKA` on all 5 cities.

## Visualization

The following code blocks below will be visualizations from the data.

```{r barplot airline delays and ontime flights}
ggplot(airlineData, aes(x = reorder(airline, -flightFrequency))) +
  geom_bar(aes(y = ontimeFlights, fill = "On-Time"), stat = "identity") +
  geom_bar(aes(y = -flightDelay, fill = "Delayed"), stat = "identity") +
  labs(title = "On-Time and Delayed Flights by Airline", x = "Airline", y = "Number of Flights", fill = "Status") +  # Change legend title
  scale_fill_manual(values = c("On-Time" = "skyblue", "Delayed" = "salmon")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

Though `AM WEST` has the highest frequencies, it appears to have the most delayed flights as opposed to `ALASKA`.

```{r Delay Rate City}
ggplot(cityairlineData, aes(x = airline, y = delayRate, fill = airline)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Delay Rate by Airline and City", 
       x = "Airline", y = "Delay Rate") +
  facet_wrap(~ city, scales = "free") +  # Facet by city
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(fill = guide_legend(title = "Airline"))

```

This bar plot shows that `AM WEST` has a higher delay rate than `ALASKA`, and reinforces the statement about Phoenix having the highest ontime rate, but lowest ontime rate, whereas San Francisco has the highest delay rate for both airlines.

## Conclusion

I believe if I had more data, such as more cities for each airline, or maybe more airlines, we can get a general understanding on what may be causing the delays for each city. I believe there is also an r library for flights that can be utilized to get a general understanding on why there are flights not arriving on time for San Francisco. I do believe if we include factors such as air traffic, weather, and other uncontrollable factors, this will give us a general understanding on *why* this is occurring.
